\emph{Gaia} DR2 identifies VFTS 682 as the most massive runaway star known to date

### 2018-05-18 VFTS682.tex
dealt with Alex comments on the text

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, J. Bestenlehner$^{5}$,
  A.~de~Koter$^{1}$, V.~H\'enault-Brunet$^{6}$, S.~Justham$^{7,8}$,
  I.~Platais$^{9}$, H.~Sana$^{}$, F.~R.~Schneider$^{10}$, J.~S.~Vink$^{11}$}

  % ,
%  A.~de~Koter$^{1}$ \todo{... add coauthors...} }

\affil{$^{}$Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands \\
    $^{}$ ESA, European Space Astronomy Centre, Apdo. de Correos 78,
    E-28691 Villanueva de la Ca\~nada, Madrid, Spain \\
      $^{}${Department of Physics and Astronomy, Hicks Building,
      Hounsfield Road, University of Sheffield, Sheffield S3 7RH, UK}\\
     $^{}$ {School of Astronomy \& Space Science, University of the Chinese
   Academy of Sciences, Beijing 100012, China}\\
    $^{}$ {National Astronomical Observatories, Chinese Academy of
      Sciences, Beijing 100012, China}\\
     $^{}$ {Department of Physics, University of Oxford, Keble Road,
       Oxford OX1 3RH, UK} \\
    $^{}$ {Space Telescope Science Institute, 3700 San Martin Drive,
       Baltimore, MD 21218, USA}\\
      {Center for Astrophysical Sciences, Department of Physics \& Astronomy, Johns Hopkins University, Baltimore, MD 21218, USA}\\
     $^{}$ {Armagh Observatory, College Hill, Armagh BT61 9DG, UK}\\
     $^{}$ {National Research Council, Herzberg Astronomy \& Astrophysics, 5071 West Saanich Road, Victoria, BC, V9E 2E7, Canada}     
}

%\thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous spectroscopic studies identified VFTS 682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 29\,pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, led to speculations whether such extreme stars can form in relative isolation. 
%
We use the unprecedented proper motions from the second \emph{Gaia} data
release (DR2) to investigate the kinematics of this object. We show
that VFTS 682 is a bonafide runaway star moving away from the star
cluster R136 with a relative
projected two-dimensional speed of about $\sim$$34\pm17\kms$ ,
implying a three-dimensional speed of about $\sim$$40\pm20\kms$ when
accounting for the velocity component along the line of sight. The
inferred age, distance, and the magnitude and direction of the
velocity are consistent with a dynamical ejection scenario from the cluster. 
%
This makes VFTS 682 the most massive runaway star known to date. It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above $100\Msun$ inside the cluster, the internal dynamics in dense star clusters and the final fate of this object.

@@

How massive stars form is one of the major longstanding questions in astrophysics
\citep[e.g.,][]{zinnecker:07}. Obtaining clues from observations has been challenging, because massive stars are intrinsically rare, 
%\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, 
evolve fast, typically reside in dense groups and they remain enshrouded in their parent cloud during the formation
process. Major progress has been made on the theoretical side,
\citep[e.g.][]{kuiper:15,rosen:16}, but simulations cannot yet resolve
the smallest length scales that are of interest to understand the 
high level of multiplicity among massive stars  \citep[e.g.,][]{bate:09, sana:17}. Understanding massive star formation, and its
possible dependence on environment and metallicity, is crucial for
understanding the role massive stars play within their host galaxies
as sources of feedback, but also for understanding the transients that
mark their death and the compact remnants they leave behind. This is
therefore a key question for the present and upcoming transient
surveys e.g., Large Synoptic Spectroscopic Survey and LIGO/Virgo,
%\citep[][]{}\todo{ref} . Blackgem is still hypothetical and a very tiny project compared to the others. 
which  will reveal transients associated to massive stars
evolution and death.

In has been proposed that most, if not all, stars form in clusters \citep{lada:03}, where massive stars are thought to  form preferentially in the dense cores. In this picture, field stars are primarily the result of the dissolution of dense groups. 
However, a significant population of massive stars exists in relative
isolation,  far from dense clusters or OB associations. The origin of
this population still remains debated \citep{lamb:16,ward:18}.  One
hypothesis is that they  formed in the field, but this poses a
challenge for the theories of star formation. The alternative
hypothesis is that these massive stars formed in clusters, and then
were ejected from their birth locations. Such ejections may result
from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems by the core-collapse of the companion star \citep[e.g.,][]{zwicky:57, blaauw:61}.

An interesting contribution to the debate on whether or not massive
stars form in relative isolation was presented by
\cite{bestenlehner:11} and \cite{bressert:12}, who discuss the case of the very massive star
VFTS 682.  This star is located in the field of the 30 Doradus region
in the Large Magellanic Cloud (LMC) and was studied as part of the
multi-epoch spectroscopic VLT-FLAMES Tarantula Survey (VFTS) of
massive stars \citep{evans:11}. Its spectral type is WNh5 and its 
present-day mass is $137.8^{+27.5}_{-15.9}\,M_\odot$,
\citep{schneider:18}. These correspond to an inferred initial mass of
$150.0^{+28.7}_{-17.4}\,M_\odot$ and a wind mass loss rate of
$\sim10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$ inferred from the
spectral analysis not accounting for clumping
\citep[][]{bestenlehner:11}, which make it one of the
most extreme objects in the region. It is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
identified by
\citet{crowther:10, crowther:16} in the core of the
R136 cluster.

Most remarkable is the location of VFTS 682, at a projected distance of
$\sim$$29$\,pc from the core of the star cluster
R136. \citet{bestenlehner:11} discuss two possibilities for the
origin of VFTS 682. Quoting directly from their abstract: "(i) the
star either formed in situ, which would have profound implications for
the formation mechanism of massive stars, or (ii) VFTS 682 is a slow
runaway star that originated from the dense cluster R136, which would
make it the most massive runaway known to date." Based on their
dynamical models, \citet{fujii:11} and \citet{banerjee:12} argued that
ejection of VFTS682 from R136 would be natural. Testing the origin of
VFTS682 is not only important for constraining isolated star
formation; it should also help to improve understanding of early star-cluster dynamics.

@@

Recently, \citet{platais:15,platais:18} used multi-epoch Hubble Space
Telescope imaging to identify and shed new light on runaway stars from
the 30 Doradus region based on proper motion measurements. In this study we report on our analysis of the data from the second \emph{Gaia} data release \cite[DR2,][]{gaia:16,brown:18}. We combine the radial velocity measurements from the VFTS survey \citep[][]{evans:11} with the proper motion from \emph{Gaia} DR2 to reconstruct the three-dimensional velocity of VFTS 682, and test the hypothesis that this star was ejected from R136.

@@

The star VFTS 682  \citep{evans:11}, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} \citep[][]{evans:11}
was originally classified as a young stellar object \citep{gruendl:09}
based on its mid-infrared excess. \citet{evans:11} classified the
object as Wolf-Rayet star of spectral type WNh5 using spectra from the
VFTS survey. The spectra available covered $\lambda$4000--7000 and
were taken at multiple epochs. Using the same dataset,
\citet{bestenlehner:11} excluded the presence of a close companion
with high confidence from the absence of radial velocity variations.

\citet{bestenlehner:11} further analyzed the spectra using the non-LTE
model atmosphere code CMFGEN \citep{hillier:98} to derive the stellar
parameters and surface abundances. They inferred an extinction
of $R_V\simeq4.7$, leading to the high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citet{schneider:18} estimated
a present-day mass for VFTS 682 of $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$, using the Bayesian analysis tool Bonnsai
\citep{schneider:17}, based on the evolutionary models from
\citet{brott:11} and \cite{kohler:15}. These values place VFTS 682 around the
``canonical upper limit'' of $\sim$$150\Msun$ by \citet{figer:05}.

@@

Further worth noticing is the variability of the
star. \citet{parker:93} reported B and V magnitudes roughly 0.5\,dex
lower than what found by \citet{evans:11}. \citet{bestenlehner:11} discuss the Optical Gravitational Lensing
Experiment (OGLE-III) light curves \citep{udalski:08} and show a
variability in the V-band at $\sim$10\% level on a timescale of years.
They commented that this is unusual for Wolf-Rayet stars and more reminiscent
of Luminous Blue Variable (LBV) stars.

We obtain the peculiar radial velocity $\delta v_\mathrm{rad}$ of VFTS 682 as the difference
between the average radial velocity of the 30 Doradus region
($270\pm10\kms$) minus the radial velocity measured from the NV $\lambda4944$
line for VFTS 682  \citep[$300\pm10\kms$, ][]{bestenlehner:11}. We implicitly assume that the use
of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not
introduce significant errors. We also note that the determination of
radial velocities for emission-lines stars is difficult because of the
presence of an optically thick, likely non-homogeneous, and variable
wind hiding the stellar surface.

@@

We define a local standard frame of reference to derive the peculiar velocity
of VFTS 682 with respect to the cluster R136 by selecting from the \emph{Gaia} DR2 catalog following closely the approach of \cite{vandermarel:02,lennon:18}.
We select all the stars in a target of 0.05 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, the errors on the proper
motion components to be smaller than 0.07\,$\mathrm{mas\
  yr^{-1}}$. At the distance to the LMC, $0.1\mathrm{mas\
  yr^{-1}}\simeq25\,\kms$ \citep[e.g.,][]{platais:18}. We
remove obvious foreground stars by requiring the absolute value of
the reported parallaxes smaller than $0.1\,\mathrm{mas\ yr^{-1}}$.
% At the distance to the
% LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
% of the proper motions removes stars that would have projected
% tangential velocities in excess of $\sim$500\,$\kms$, which are most
% likely to be foreground stars. We checked that the additional
% requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
% not reduces further our sample.
This selection yields a sample of 66 stars.% ,

@@

@@

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec}^{Gaia} = \sqrt{\left(\delta v_\mathrm{RA}^{Gaia}\right)^2
    +\left(\delta v_\mathrm{DEC}^{Gaia}\right)^2+\left(\delta
      v_\mathrm{rad}^{VFTS}\right)^2} = 46 \pm 18 
  \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS 682 with respect the
surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

\section{Validation with HST proper motions}
\todo{maybe rename to Astrometry from HST and make sec. 2.1.1 or 2.2
  but above Gaia data?}

Recently, \citet{platais:18} presented Hubble Space Telescope (HST)  astrometry and proper
motions in the 30 Doradus region based on two epochs of
observations; the first epoch (GO-12499; P.I.: D.~J.~Lennon)
in October 3-8 2011 and the second epoch (GO-13359) in
October 6-11 2014, both using the F775W filter.
The second epoch was designed to match the pointing and
orientation of the first epoch as exactly as possible so
that the displacement measurements could be purely differential.
While most bright massive stars (V$<$14) were saturated in
their data and missing from that catalog, due to its
high extinction, E(B-V)=0.9, this star is faint enough
to be included in that catalog. Its proper motions are
listed in \Tabref{tab:vfts682} \todo{add}. For
large distances ($\gtrsim50$\,kpc), HST data can compete or even
exceed Gaia data in quality.

The catalog from \citet{platais:18} lists the proper motion
components of VFTS682 relative to the field, providing a
measure of the motion of this star completely independent from
ours. They found proper
motion components relative to the field
$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and
$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
  yr^{-1}}$.

\todo{redo below with average HST+Gaia?Actually redo also calculation
  of $v_\mathrm{pec}$}
\subsection{Does it come from the R136 cluster?}
\label{sec:r136_origin}

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} \simeq
  \frac{29\,\mathrm{pc}}{34\,\kms} \simeq 0.85\pm\,0.02\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS 682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =34\pm
1\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}.% , and we use the approximation $1 \kms \simeq 1
% \mathrm{pc \ Myr^{-1}}$.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates.

The kinematic age $\tau_\mathrm{kin}$ is smaller than the apparent age
of the star of $1.0\pm 0.2$\,Myr from \cite{schneider:18}, which
corroborates the idea that the star is the result of a dynamical
ejection.

@@

Dynamical ejections due to N-body interactions typically (although, not necessarily) eject the least
massive star among those interacting \cite[e.g.,][]{banerjee:12}. This means that, just
based on the kinematic properties of VFTS 682, we would expect several
stars with initial masses larger than $\sim$$150\,M_\odot$ in the
cluster R136.
This is consistent with the detection
of extremely massive stars by \cite{crowther:10} in the core of the
cluster. The projected rotational equatorial
velocity\footnote{However, the determination of the rotational
  velocity for stars showing lines in emission is complicated by the optically thick wind screening the surface of the star, and should be
  considered with caution.} of VFTS 682
reported by \cite{schneider:18} is $v\sin(i)<200\,\kms$, which is in
line with the average rotation rate of massive stars in the region
\citep[][]{ramirez-agudelo:15}. This suggests that VFTS 682 (i) has not
experienced rotationally induced chemically homogeneous evolution
\citep[][]{maeder:00,demink:09,marchant:16}, and (ii) it has not
accreted mass from or merged with a binary companion, nor will it, since the multi-epoch
data of the VFTS survey rule out the presence of a close companion at the
present day. Moreover, the spectral type of VFTS 682
\citep[WNh5,][]{bestenlehner:11} is the same as R136a1-a3, i.e.~the
three most massive stars detected in the core of the cluster by
\cite{crowther:10}, with an astonishing similarity in particular with
the spectrum of R136a3. Therefore,
VFTS 682 might be an ideal target to constrain the stellar physics of
stars with masses well above $\sim$$100\,M_\odot$: its isolation makes
it an easier target for observations compared to the similar stars
present in the crowded core of R136.

\citet{banerjee:12} used N-body simulations of fully segregated
clusters with all massive stars in binaries to suggest that VFTS 682
was ejected from R136, in agreement with our findings. They
demonstrated that the cluster potential does not significantly change
the velocity of the star after the ejection. In their
model, they relied on (dynamically driven) stellar mergers to explain the high masses of
VFTS 682 and the massive members of R136. While it is not possible to
robustly exclude that VFTS 682 is itself a merger, as already
mentioned, its spectrum does not show any of the obvious signatures
like fast rotation. To eject such massive object, the cluster must have produced
a large number of massive runaways. Indeed, several relatively
isolated massive stars are observed in the region, some with known
large radial velocities. A comprehensive study of the kinematic
properties of all the massive stars surrounding R136 might shed light
on whether some can be unequivocally identified as merger products. It
is also possible that the star or binary that caused the ejection of
VFTS 682 might have been ejected in the opposite direction, and is also
in isolation at present day.

@@

The numerical simulations from \cite{oh:16} suggest that dynamical
interaction eject the majority of the stars during of shortly after the cluster
core-collapse. The large number of isolated massive stars around it
suggest that R136 has already evolved past the
time of maximum stellar density. This might have implications for the
question on whether the cluster formed via a monolithic collapse, or
as a (potentially ongoing) merger of several sub-clusters.

>>>>

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, I.~Platais$^{5}$, J. Bestenlehner$^{6}$,
  A.~de~Koter$^{1}$, V.~H\'enault-Brunet$^{7}$, S.~Justham$^{8}$, H.~Sana$^{9}$, F.~R.~N.~Schneider$^{10}$, J.~S.~Vink$^{11}$}

\affil{\todo{check affil. num}$^{}$Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands \\
    $^{}$ ESA, European Space Astronomy Centre, Apdo. de Correos 78,
    E-28691 Villanueva de la Ca\~nada, Madrid, Spain \\
      $^{}${Department of Physics and Astronomy, Hicks Building,
      Hounsfield Road, University of Sheffield, Sheffield S3 7RH, UK}\\
     $^{}$ {School of Astronomy \& Space Science, University of the Chinese
   Academy of Sciences, Beijing 100012, China}\\
    $^{}$ {National Astronomical Observatories, Chinese Academy of
      Sciences, Beijing 100012, China}\\
     $^{}$ {Department of Physics, University of Oxford, Keble Road,
       Oxford OX1 3RH, UK} \\
    $^{}$ {Space Telescope Science Institute, 3700 San Martin Drive,
       Baltimore, MD 21218, USA}\\
      {Center for Astrophysical Sciences, Department of Physics \& Astronomy, Johns Hopkins University, Baltimore, MD 21218, USA}\\
     $^{}$ {Armagh Observatory, College Hill, Armagh BT61 9DG, UK}\\
     $^{}$ {National Research Council, Herzberg Astronomy \& Astrophysics, 5071 West Saanich Road, Victoria, BC, V9E 2E7, Canada}     
}

%\thanks{Corresponding author:  M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
The formation and fates of the most massive stars remains among the most elusive questions in stellar astrophysics. 
%
Previous spectroscopic studies identified VFTS 682 as one of the most
massive stars known. It resides in the field of the 30 Dor region in
the Large Magellanic Cloud at a projected distance of 29\,pc of the
star cluster Radcliffe 136 (R136). The lack of a detected offset in radial velocity\todo{can we really say that?RV
  is about 30kms...}
derived from the spectra, led to speculations whether such extreme
stars can form in relative isolation . 
%
\todo{add mention of HST}We use the unprecedented proper motions from the second \emph{Gaia} data
release (DR2) to investigate the kinematics of this object. We show
that VFTS 682 is a bonafide runaway star moving away from the star
cluster R136 with a relative
projected two-dimensional speed of about $\sim$$34\pm17\kms$ ,
implying a three-dimensional speed of about $\sim$$40\pm20\kms$ when
accounting for the velocity component along the line of sight. The
inferred age, distance, and the magnitude and direction of the
velocity are consistent with a dynamical ejection scenario from the cluster. 
%
This makes VFTS 682 the most massive runaway star known to date. It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above $100\Msun$ inside the cluster, the internal dynamics in dense star clusters and the final fate of this object.

@@

How massive stars form is one of the major longstanding questions in astrophysics
\citep[e.g.,][]{zinnecker:07}. Obtaining clues from observations has been challenging, because massive stars are intrinsically rare, 
%\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, 
evolve fast, typically reside in dense groups and remain enshrouded in
their parent cloud during the entirety of their formation
process. Major progress has been made on the theoretical side,
\citep[e.g.][]{kuiper:15,rosen:16}, but simulations cannot yet resolve
the smallest length scales that are of interest to understand the 
high level of multiplicity among massive stars  \citep[e.g.,][]{bate:09, sana:17}. Understanding massive star formation, and its
possible dependence on environment and metallicity, is crucial for
understanding the role massive stars play within their host galaxies
as sources of feedback, but also for understanding the transients that
mark their death and the compact remnants they leave behind. This is
therefore a key question for the present and upcoming transient
surveys, e.g., Large Synoptic Spectroscopic Survey and LIGO/Virgo,
%\citep[][]{}\todo{ref} . Blackgem is still hypothetical and a very tiny project compared to the others. 
which  will reveal transients associated to massive stars
evolution and death.

In has been proposed that most, if not all, stars form in clusters \citep{lada:03}, where massive stars are thought to  form preferentially in the dense cores. In this picture, field stars are primarily the result of the dissolution of dense groups. 
However, a significant population of massive stars exists in relative
isolation,  far from dense clusters or OB associations. The origin of
this population still remains debated \citep{gvaramadze:12, lamb:16,ward:18}.  One
hypothesis is that they  formed in the field, but this poses a
challenge for the theories of star formation. The alternative
hypothesis is that these massive stars formed in clusters, and then
were ejected from their birth locations. Such ejections may result
from dynamical interactions \citep[e.g.,][]{poveda:67} or from the
disruption of binary systems by the core-collapse of the companion
star \citep[e.g.,][]{zwicky:57, blaauw:61, renzo:18}.

An interesting contribution to the debate on whether or not massive
stars form in relative isolation was presented by
\cite{bestenlehner:11} and \cite{bressert:12}, who discussed the case of the very massive star
VFTS 682.  This star is located in the field of the 30 Doradus region
in the Large Magellanic Cloud (LMC) and was studied as part of the
multi-epoch spectroscopic VLT-FLAMES Tarantula Survey (VFTS) of
massive stars \citep{evans:11}. Its spectral type is WNh5 and its 
present-day mass is $137.8^{+27.5}_{-15.9}\,M_\odot$,
\citep{schneider:18}. These correspond to an inferred initial mass of
$150.0^{+28.7}_{-17.4}\,M_\odot$. It features a wind mass loss rate of
$\sim10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, inferred from
spectral analysis assuming a smooth outflow, i.e., not accounting for clumping
\citep[][]{bestenlehner:11}. These properties make it one of the
most extreme objects in the region. It is reminiscent of the very
massive stars % , a.k.a. ``monster stars'',
identified by
\citet{crowther:10, crowther:16} in the core of the
R136 cluster.

Most remarkable is the location of VFTS 682, at a projected distance of
$\sim$$29$\,pc from the core of the star cluster
R136. \citet{bestenlehner:11} discuss two possibilities for the
origin of VFTS 682. Quoting directly from their abstract: "(i) the
star either formed in situ, which would have profound implications for
the formation mechanism of massive stars, or (ii) VFTS 682 is a slow
runaway star that originated from the dense cluster R136, which would
make it the most massive runaway known to date". Based on their
dynamical models, \citet{fujii:11} and \citet{banerjee:12} argued that
ejection of VFTS 682 from R136 would be natural. Testing the origin of
VFTS 682 is not only important for constraining isolated star
formation; it should also help to improve understanding of early star-cluster dynamics.

@@

Recently, \citet{platais:15,platais:18} used multi-epoch Hubble Space
Telescope (HST) imaging to identify and shed new light on runaway stars from
the 30 Doradus region based on proper motion
measurements. \citet{lennon:18} used data from the second \emph{Gaia}
data release \cite[DR2,][]{gaia:16,brown:18} to further investigate
the isolated O-type stars in the vicinity of R136. In this study, we
report on our analysis of the \emph{Gaia} DR2 data for VFTS 682 which
was not part of the sample of \citet{lennon:18} because of its WNh5
spectral type. We combine the radial velocity measurements from the
VFTS survey \citep[][]{evans:11} with the proper motion from
\emph{Gaia} DR2 to reconstruct the three-dimensional velocity of VFTS
682, and test the hypothesis that this star was ejected from R136. We
use the HST proper motions from \citet{platais:18} to validate our
analysis of the \emph{Gaia} DR2 data.

@@

The star VFTS 682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} \citep[][see also
\Tabref{tab:vfts682} for the coordinates from \emph{Gaia} DR2 in the
ICGS frame]{evans:11}
was originally classified as a young stellar object \citep{gruendl:09}
based on its mid-infrared excess. \citet{evans:11} classified the
object as Wolf-Rayet star of spectral type WNh5 using spectra from the
VFTS survey. The spectra available covered $\lambda$4000--7000 and
were taken at multiple epochs. Using the same dataset,
\citet{bestenlehner:11} excluded the presence of a close companion
with high confidence from the absence of radial velocity variations.

\citet{bestenlehner:11} further analyzed the spectra using the non-LTE
model atmosphere code CMFGEN \citep{hillier:98} to derive the stellar
parameters and surface abundances. They inferred an extinction
of $A_V=4.45\pm0.12$, leading to the high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citet{schneider:18} estimated
a present-day mass for VFTS 682 of $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$, using the Bayesian analysis tool BONNSAI
\citep{schneider:17}, based on the evolutionary models from
\citet{brott:11} and \cite{kohler:15}. % These values place VFTS 682 around the
% ``canonical upper limit'' of $\sim$$150\Msun$ by \citet{figer:05}.

@@

Further worth noticing is the variability of the
star. \citet{parker:93} reported B and V magnitudes roughly 0.5\,magnitude
lower than what found by \citet{evans:11}. \citet{bestenlehner:11} discuss the Optical Gravitational Lensing
Experiment (OGLE-III) light curves \citep{udalski:08} and show a
variability in the V-band at $\sim$10\% level on a timescale of years.
They commented that this is unusual for Wolf-Rayet stars and more reminiscent
of Luminous Blue Variable (LBV) stars.

We obtain the peculiar radial velocity $\delta v_\mathrm{rad}$ of VFTS 682 as the difference
between the average radial velocity of the 30 Doradus region
($270\pm10\kms$) minus the radial velocity measured from the NV $\lambda4944$
line for VFTS 682  \citep[$300\pm10\kms$, ][]{bestenlehner:11}. We implicitly assume that the use
of a slightly different reference frame for $\delta v_\mathrm{rad}$ does not
introduce significant errors. We also note that the determination of
radial velocities for emission-lines stars is difficult because of the
presence of an optically thick, likely non-homogeneous, and variable
wind obscuring the stellar surface.

\subsection{HST astrometry for VFTS682}
Recently, \citet{platais:18} presented HST  astrometry and proper
motions in the 30 Doradus region based on two epochs of
observations; the first epoch (GO-12499; P.I.: D.~J.~Lennon)
in October 3-8 2011 and the second epoch (GO-13359) in
October 6-11 2014, both using the F775W filter.
The second epoch was designed to match the pointing and
orientation of the first epoch as exactly as possible so
that the displacement measurements could be purely differential.
While most bright massive stars (V$<$14) were saturated in
their data and missing from that catalog, due to its
high extinction, E(B-V)=0.9, this star is faint enough
to be included in that catalog. Its proper motions are
listed in \Tabref{tab:vfts682}. For
large distances ($\gtrsim50$\,kpc), HST astrometry can compete or even
exceed \emph{Gaia} data in quality.

The catalog from \citet{platais:18} lists the proper motion
components of VFTS 682 relative to the field, providing a
measure of the motion of this star completely independent from
ours. They found proper
motion components relative to the field
$\delta\mu_\mathrm{RA}^\mathrm{HST} = 0.01\pm0.13\,\mathrm{mas\
  yr^{-1}}$ and
$\delta\mu_\mathrm{DEC}^\mathrm{HST}=0.2\pm0.1\,\mathrm{mas\
  yr^{-1}}$.

\todo{redo below with average HST+Gaia?Actually redo also calculation
  of $v_\mathrm{pec}$}

@@

We define a local standard frame of reference to derive the peculiar velocity
of VFTS 682 with respect to the cluster R136 by selecting from the \emph{Gaia} DR2 catalog following closely the approach of \cite{vandermarel:02,lennon:18}.
We select all the stars in a target of 0.05 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require a G-band
magnitude brighter than 17, corresponding roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, and the errors on the proper
motion components to be smaller than 0.07\,$\mathrm{mas\
  yr^{-1}}$. At the distance to the LMC, $0.1\mathrm{mas\
  yr^{-1}}\simeq25\,\kms$ \citep[e.g.,][]{platais:18}. We
remove obvious foreground stars by requiring the absolute value of
the reported parallaxes smaller than $0.1\,\mathrm{mas\ yr^{-1}}$.
% At the distance to the
% LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
% of the proper motions removes stars that would have projected
% tangential velocities in excess of $\sim$500\,$\kms$, which are most
% likely to be foreground stars. We checked that the additional
% requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
% not reduces further our sample.
This selection yields a sample of 66 stars.% ,

@@

\todo{nebular lines to compare velocity of the gas?}
\todo{fix numbers}

@@

\begin{equation}
  \label{eq:speed_around}
  v_\mathrm{pec}^{Gaia} = \sqrt{\left(\delta v_\mathrm{RA}^{Gaia}\right)^2
    +\left(\delta v_\mathrm{DEC}^{Gaia}\right)^2+\left(\delta
      v_\mathrm{rad}^{VFTS}\right)^2} = 46 \pm 18 
  \kms \ .
\end{equation}
This value for the three-dimensional speed of VFTS 682 with respect to
the surrounding stars make it the most massive ``bona fide'' runaway star
known to date.

\subsection{Does it come from the R136 cluster?}
\label{sec:r136_origin}

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} \simeq
  \frac{29\,\mathrm{pc}}{34\,\kms} \simeq 0.85\pm\,0.02\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS 682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =34\pm
1\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}.% , and we use the approximation $1 \kms \simeq 1
% \mathrm{pc \ Myr^{-1}}$.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, which amounts to about
$\sim$$2.2\%$, and is negligible compared to other uncertainties.

The kinematic age $\tau_\mathrm{kin}$ is similar or smaller than the apparent age
of the star of $1.0\pm 0.2$\,Myr from \cite{schneider:18}, which
corroborates the idea that the star is the result of a dynamical
ejection.

@@

Dynamical ejections due to N-body interactions typically (although, not necessarily) eject the least
massive star among those interacting \cite[e.g.,][]{banerjee:12}. This means that, just
based on the kinematic properties of VFTS 682, we would expect several
stars with initial masses larger than $\sim$$150\,M_\odot$ in the
cluster R136.
This is consistent with the detection
of extremely massive stars by \cite{crowther:10} in the core of the
cluster. The projected rotational equatorial
velocity\footnote{However, the determination of the rotational
  velocity for stars showing lines in emission is complicated by the optically thick wind screening the surface of the star, and should be
  considered with caution.} of VFTS 682
reported by \cite{schneider:18} is $v\sin(i)<200\,\kms$, which is in
line with the average rotation rate of massive stars in the region
\citep[][]{ramirez-agudelo:15}. This suggests that VFTS 682 (i) has not
experienced rotationally induced chemically homogeneous evolution
\citep[][]{maeder:00,demink:09,marchant:16}, and (ii) it has not
accreted mass from or merged with a binary companion, nor will it, since the multi-epoch
data of the VFTS survey rule out the presence of a close companion at the
present day. Moreover, the spectral type of VFTS 682
\citep[WNh5,][]{bestenlehner:11} is the same as R136a1-a3, i.e.~the
three most massive stars detected in the core of the cluster by
\cite{dekoter:97,crowther:10,crowther:16}, with an astonishing similarity in particular with
the spectrum of R136a3. Therefore,
VFTS 682 might be an ideal target to constrain the stellar physics of
stars with masses well above $\sim$$100\,M_\odot$: its isolation makes
it an easier target for observations compared to the similar stars
present in the crowded core of R136.

\citet{banerjee:12} used N-body simulations of fully segregated
clusters with all massive stars in binaries to suggest that VFTS 682
was ejected from R136. They
demonstrated that the cluster potential does not significantly change
the velocity of the star after the ejection. In their
model, they relied on (dynamically driven) stellar mergers to explain the high masses of
VFTS 682 and the massive members of R136. While it is not possible to
robustly exclude that VFTS 682 is itself a merger, as already
mentioned, its spectrum does not show any of the obvious signatures
like fast rotation. To eject such a massive object, the cluster is
expected to have produced
a large number of massive runaways. Indeed, several relatively
isolated massive stars are observed in the region, some with known
large radial velocities. A comprehensive study of the kinematic
properties of all the massive stars surrounding R136 might shed light
on whether some can be unequivocally identified as merger products. It
is also possible that the star or binary that caused the ejection of
VFTS 682 might have been ejected in the opposite direction, and is also
in isolation at present day.

@@

The numerical simulations from \cite{oh:16} suggest that dynamical
interaction eject the majority of the stars during or shortly after the cluster
core-collapse. The large number of isolated massive stars around it
suggest that R136 has already evolved past the
time of maximum stellar density. This might have implications for the
question on whether the cluster formed via a monolithic collapse, or
as a (potentially ongoing) merger of several sub-clusters.
